---
title: "Classification analysis of Acute Respiratory Distress Syndrome (ARDS) using preECMO data"
author: "Kavya K H"
output:
  html_document:
    toc: TRUE
    toc_float: true
  pdf_document: 
      extra_dependencies: ["float"]
date: "15 June"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts=list(width.cutoff=80),tidy=TRUE)
```

```{r rlibraries, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
# Load required packages
library(readxl)
library(tidyverse)
library(ggplot2)
library(corrplot)
library(gridExtra)
library(dplyr)
library(glmnet)
library(caret)
library(naniar)
library(pROC)
library(mltools)

# Read dataset
data <- read_excel("ARDSdata.xlsx")
```

```{r name, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}

```
## Data description
The dataset contain medical records of 450 patients; a full list of variable description is given below.

```{r variables , echo=FALSE, results='asis'}
cat(' Table: \\label{tab:variables}Description of Variables:
  
| Variable| Description|
|-----|----------|
|Pt_ID |Patient ID|
|Gender| Gender|
|Age |Age in number of days|
|Indication |acute lung failure (ALF), viral pneumonia (1), bacterial pneumonia (2), aspiration pneumonitis (3), ARDS Trauma (4), ARDS surgery (5), Chemo (6), Other (7)|
|ECMO_Survival |Survival indicator|
|RR|Respiratory rate|
|Vt |Tidal volume|
|FiO2 |Inspired fraction of oxygen|
|Ppeak |Peak airway pressure|
|Pmean |Mean airway pressure|
|PEEP |Positive end expiratory pressure|
|PF |Arterial partial pressure of oxygen|
|SpO2 |Periperal oxygen saturation|
|PaCO2 |Arterial partial pressure of carbon dioxide|
|pH |Arterial pH|
|BE | Arterial base excess|
|Lactate |Arterial lactate|
|NAdose |Noradrenaline dose|
|MAP |Mean arterial pressure|
|CK |Creatinine Kinase|
|CRP |C reactive protein|
|ATIII |Anti-thrombin III|
|HB |Haemaglobin|
|Creatinine, Fibrinogen, Bilirubin, Albumin, | Urea, Ddimer, Leukocytes, Platelets, TNFa, IL6, IL8, siL2|
|||
')
```

# Data Exploration

```{r explore, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
# Remove day1 ECMO columns and convert categorical variables as factors
preECMO <- data %>% 
  select(-starts_with("Day1")) %>%
  mutate(Indication = as.factor(Indication),
         ECMO_Survival = as.factor(ECMO_Survival),
         Hospital_Survival = as.factor(Hospital_Survival),
         Gender = as.factor(Gender),
         Pt_ID = as.character(Pt_ID))

# check for patient repetitions
paste("Repetition of patient records in the data: ",length(unique(preECMO$Pt_ID)) != nrow(preECMO))

# View data
str(preECMO)

# Remove 'PreECMO_' from variables names
names(preECMO) = gsub(pattern = "PreECMO_", replacement = "", x = names(preECMO))

# Shuffle data
shuffle_index <- sample(1:nrow(preECMO))
preECMO <- preECMO[shuffle_index, ]

# View data summary
summary(preECMO)

# check number of missing values per patient
preECMO <- preECMO %>%
  mutate(naPatient = rowSums(is.na(preECMO)))

# save continuous variables into a dataframe
ndata1 <- preECMO %>%
  select(-c(Pt_ID, ECMO_Survival, Hospital_Survival, Gender, Indication, naPatient))

# change the names of levels in response variable for clarity
preECMO$ECMO_Survival = as.factor(ifelse(preECMO$ECMO_Survival == "Y", "Survived", "Died"))

```

## Data Visualisation

### Barplots
```{r plots1, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE, fig.align = "center"}
# Barplot for the variable ECMO_Survival
ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = ..count..)) +
  geom_bar() +
  xlab("ECMO Survival") +
  ylab("Proportion") + ggtitle("Proportion of ECMO survival indicator")

# Barplot for the variable Gender
ggplot(preECMO, mapping = aes(x = Gender, y = ..count.., fill = ECMO_Survival)) +
  geom_bar() +
  xlab("Gender") +
  ylab("Proportion") + ggtitle("Proportion of Gender")

# Barplot for the variable Indication
ggplot(preECMO, mapping = aes(x = Indication, y = ..count.., fill = ECMO_Survival)) +
  geom_bar() +
  xlab("") +
  ylab("Proportion")+
  scale_x_discrete(labels=c("viral pneumonia", "bacterial pneumonia", 
                            "aspiration pneumonitis", "ARDS Trauma", "ARDS surgery",
                            "Chemo", "Other")) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("Proportion of different disease indicators in the dataset")

# piechart for the variable Indication
preECMO %>% 
  ggplot(aes(x = "", fill = Indication)) + 
  geom_bar(position = "fill", width = 1) + 
  coord_polar(theta = "y") + 
  labs( 
    title = "Disease Indicator proportions",
    x = "", 
    y = ""
  ) + 
  scale_fill_brewer(palette="Blues")+
  theme(
    panel.grid = element_blank(), 
    axis.ticks = element_blank(), 
    axis.text.x=element_blank(), 
    panel.border = element_blank() 
    ) +
  facet_grid(. ~ ECMO_Survival)

# preECMO$ECMO_Survival = as.factor(ifelse(preECMO$ECMO_Survival == "Survived", "Y", "N"))
```

### Boxplots 
```{r plots2, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE, fig.align = "center"}
# -----------------------------------------------------------------------------
# boxplots
My_Theme = theme(plot.title = element_text(size=11),
                 axis.text=element_text(size=10),
                 axis.title=element_text(size=10),
                 legend.title = element_text(size=10),
                 legend.text = element_text(size=9))

p1 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = Duration_ECMO)) +
  geom_boxplot(aes(fill = ECMO_Survival) )+
  labs(x = "Survival", y = "Duration") + My_Theme

p2 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = RR)) +
  geom_boxplot(aes(fill = ECMO_Survival)) +
  labs(x = "Survival", y = "RR") + My_Theme

p3 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = Vt)) +
  geom_boxplot(aes(fill = ECMO_Survival)) +
  labs(x = "Survival", y = "Vt") + My_Theme

p4 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = FiO2)) +
  geom_boxplot(aes(fill = ECMO_Survival)) +
  labs(x = "Survival", y = "FiO2") + My_Theme

p5 <-  ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = Pmean)) +
  geom_boxplot(aes(fill = ECMO_Survival)) +
  labs(x = "Survival", y = "Pmean") + My_Theme

p6 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = SpO2)) +
  geom_boxplot(aes(fill = ECMO_Survival)) +
  labs(x = "Survival", y = "SpO2") + My_Theme

grid.arrange(p1, p2, p3, p4, p5, p6)

p11 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = PaCO2)) +
  geom_boxplot(aes(fill = ECMO_Survival)) +
  labs(x = "Survival", y = "PaCO2") + My_Theme

p12 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = pH)) +
  geom_boxplot(aes(fill = ECMO_Survival)) +
  labs(x = "Survival", y = "pH") + My_Theme

p13 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = BE)) +
  geom_boxplot(aes(fill = ECMO_Survival)) +
  labs(x = "Survival", y = "BE") + My_Theme

p14 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = Lactate)) +
  geom_boxplot(aes(fill = ECMO_Survival) )+
  labs(x = "Survival", y = "Lactate") + My_Theme

p15 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = NAdose)) +
  geom_boxplot(aes(fill = ECMO_Survival)) +
  labs(x = "Survival", y = "NAdose") + My_Theme

p16 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = MAP)) +
  geom_boxplot(aes(fill = ECMO_Survival)) +
  labs(x = "Survival", y = "MAP") + My_Theme

grid.arrange(p11, p12, p13, p14, p15, p16)

p21 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = Creatinine)) +
  geom_boxplot(aes(fill = ECMO_Survival)) +
  labs(x = "Survival", y = "creatinine") + My_Theme

p22 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = CK)) +
  geom_boxplot(aes(fill = ECMO_Survival)) +
  labs(x = "Survival", y = "Ck") + My_Theme

p23 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = Bilirubin)) +
  geom_boxplot(aes(fill = ECMO_Survival) )+
  labs(x = "Survival", y = "Bilirubin") + My_Theme

p24 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = CRP)) +
  geom_boxplot(aes(fill = ECMO_Survival)) +
  labs(x = "Survival", y = "CRP") + My_Theme

p25 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = Fibrinogen)) +
  geom_boxplot(aes(fill = ECMO_Survival)) +
  labs(x = "Survival", y = "Fibrinogen") + My_Theme

p26 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = Ddimer)) +
  geom_boxplot(aes(fill = ECMO_Survival)) +
  labs(x = "Survival", y = "Ddimer") + My_Theme

grid.arrange(p21, p22, p23, p24, p25, p26)

p31 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = ATIII)) +
  geom_boxplot(aes(fill = ECMO_Survival)) +
  labs(x = "Survival", y = "ATIII") + My_Theme

p32 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = Leukocytes)) +
  geom_boxplot(aes(fill = ECMO_Survival) )+
  labs(x = "Survival", y = "leucocytes") + My_Theme

p33 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = Platelets)) +
  geom_boxplot(aes(fill = ECMO_Survival)) +
  labs(x = "Survival", y = "Platelets") + My_Theme

p34 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = TNFa)) +
  geom_boxplot(aes(fill = ECMO_Survival)) +
  labs(x = "Survival", y = "TNFa") + My_Theme

p35 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = IL6)) +
  geom_boxplot(aes(fill = ECMO_Survival)) +
  labs(x = "Survival", y = "IL6") + My_Theme

p36 <- ggplot(preECMO, mapping = aes(x = ECMO_Survival, y = siIL2)) +
  geom_boxplot(aes(fill = ECMO_Survival)) +
  labs(x = "Survival", y = "siIL2") + My_Theme

grid.arrange(p31, p32, p33, p34, p35, p36)

```

# Hypothesis tests 

Initial feature filtering was done using simple hypothesis tests. We used shapiro-wilks test to check normality, and spearman correlation coefficient and pearson correlation coefficient for variable selection.

```{r normalitytest, eval=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
preECMO_2 <- preECMO
# Shapiro–Wilk test
# normality 
shapiro.test(preECMO$RR)
shapiro.test(preECMO$Vt)
shapiro.test(preECMO$CK)
shapiro.test(preECMO$Bilirubin)
shapiro.test(preECMO$CRP)
shapiro.test(preECMO$Fibrinogen)
shapiro.test(preECMO$Ddimer)
shapiro.test(preECMO$ATIII)
shapiro.test(preECMO$TNFa)
shapiro.test(preECMO$siIL2)
```

## Correlation 
```{r plots3, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE, fig.align = "center"}
# pearson correlation coefficient for continuous data
corr_pearson <- cor(ndata1, use = "complete.obs")
# continuous variables correlation plot 
corrplot(corr_pearson, method = "circle", tl.cex=0.5, tl.col = "black")

# Spearman correlation coefficient for continuous data
# to use kendall correlation coefficient use 'method="kendall"'
corr_spearsman <- cor(ndata1, method = "spearman", use = "complete.obs")
# continuous variables correlation plot
corrplot(corr_spearsman, method = "circle", tl.cex=0.5, tl.col = "black")

# find highly correlated variables
w <- which(abs(corr_spearsman)>0.7 & row(corr_spearsman)<col(corr_spearsman), arr.ind=TRUE)

# reconstruct names from positions
high_cor <- matrix(colnames(corr_spearsman)[w],ncol=2)
high_cor

# devtools::install_github("laresbernardo/lares")
library(lares) # library contain tool to display top correlations in the dataset

# Display top 5 highly correlated variables
corr_cross(ndata1, # name of dataset
  max_pvalue = 0.05, method = "spearman",# display only significant correlations (at 5% level)
  top = 5 # display top 5 couples of variables (by correlation coefficient)
)
# corr_cross(preECMO, type=2)

# Remove highly correlated variables (only keep one based on the number of missing values) and unused variables
preECMO <- preECMO %>%
  select(-c(PEEP, PF, IL8, naPatient, Hospital_Survival, Duration_ECMO))

# Display summary of dataset
# summary(preECMO)

```

```{r ohet, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
# One hot encoding to binarize the catagorical variable "Indication"
preECMO <- cbind(preECMO, model.matrix(~Indication - 1, data = preECMO))

# Remove the variable "Indication" from the dataset
preECMO <- preECMO %>% 
  select(-c(Indication)) 

# Convert variable types to factors
preECMO <- preECMO %>%
  mutate(Indication1 = as.factor(Indication1)) %>%
  mutate(Indication2 = as.factor(Indication2)) %>%
  mutate(Indication3 = as.factor(Indication3)) %>%
  mutate(Indication4 = as.factor(Indication4)) %>%
  mutate(Indication5 = as.factor(Indication5)) %>%
  mutate(Indication6 = as.factor(Indication6)) %>%
  mutate(Indication7 = as.factor(Indication7)) 
```

```{r chisq, echo=FALSE, eval=FALSE, warning=FALSE, message=FALSE}
# Chi square test
chisq.test(preECMO$Indication1,preECMO$ECMO_Survival,correct=F)
chisq.test(preECMO$Indication2,preECMO$ECMO_Survival,correct=F)
chisq.test(preECMO$Indication3,preECMO$ECMO_Survival,correct=F)
chisq.test(preECMO$Indication4,preECMO$ECMO_Survival,correct=F)
chisq.test(preECMO$Indication5,preECMO$ECMO_Survival,correct=F)
chisq.test(preECMO$Indication6,preECMO$ECMO_Survival,correct=F)
chisq.test(preECMO$Indication7,preECMO$ECMO_Survival,correct=F)
chisq.test(preECMO$Gender,preECMO$ECMO_Survival,correct=F)

```

# Data preprocessing
## Missing values 

```{r missing value, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
# Save dataset in a new dataframe
preECMO_original <- preECMO

# Display missing values
vis_miss(preECMO) # Display position and percentage of missing values in the dataset
preECMO %>% 
  gg_miss_var(show_pct = TRUE, facet = ECMO_Survival) # Display percentage of missing values grouped by outcome variable 
gg_miss_var(preECMO, show_pct = TRUE) # Display percentage of missing values in the dataset


# Remove the variable "Albumin" because it contain almost 50% missing values
preECMO <- preECMO %>% 
  select(-c(Albumin)) 

# Replace missing values with median (variables with percentage of missing values less than 2%)
preECMO$Lactate[is.na(preECMO$Lactate)] <- median(preECMO$Lactate, na.rm = TRUE)
preECMO$Leukocytes[is.na(preECMO$Leukocytes)] <- median(preECMO$Leukocytes, na.rm = TRUE)


# replace missing values with random values from the corresponding column (if relatively large % of NAs and the % of missing values in the column between 2 and 4)
preECMO <- preECMO %>% 
  mutate(naPatient = rowSums(is.na(preECMO))) %>% # calculate no. of NAs in each row
  filter(naPatient < 3) %>% # keep only rows in which no. of missing values less than 3
  drop_na(ATIII) %>% # drop rows that contain NAs for the variable ATIII
  mutate(CK= ifelse(!is.na(CK),CK, sample(na.omit(preECMO$CK)))) %>%
  mutate(Bilirubin= ifelse(!is.na(Bilirubin),Bilirubin, sample(na.omit(preECMO$Bilirubin)))) %>%
  mutate(CRP= ifelse(!is.na(CRP),CRP, sample(na.omit(preECMO$CRP)))) %>%
  mutate(Ddimer= ifelse(!is.na(Ddimer),Ddimer, sample(na.omit(preECMO$Ddimer)))) %>%
  mutate(Urea= ifelse(!is.na(Urea),Urea, sample(na.omit(preECMO$Urea)))) %>%
  mutate(TNFa= ifelse(!is.na(TNFa),TNFa, sample(na.omit(preECMO$TNFa)))) %>%
  mutate(siIL2= ifelse(!is.na(siIL2),siIL2, sample(na.omit(preECMO$siIL2))))

# Replace remaining NAs with median (variables with percentage of missing values less than 2%)
preECMO$RR[is.na(preECMO$RR)] <- median(preECMO$RR, na.rm = TRUE)
preECMO$Vt[is.na(preECMO$Vt)] <- median(preECMO$Vt, na.rm = TRUE)
preECMO$Ppeak[is.na(preECMO$Ppeak)] <- median(preECMO$Ppeak, na.rm = TRUE)
preECMO$Pmean[is.na(preECMO$Pmean)] <- median(preECMO$Pmean, na.rm = TRUE)
preECMO$MAP[is.na(preECMO$MAP)] <- median(preECMO$MAP, na.rm = TRUE)
preECMO$Fibrinogen[is.na(preECMO$Fibrinogen)] <- median(preECMO$Fibrinogen, na.rm = TRUE)

# Remove unwanted variables
preECMO <- preECMO %>% 
  select(-c(Pt_ID, naPatient)) 

# display summary of dataset
summary(preECMO)
```

# Data splitting
```{r lrpp, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
# concert the type of the variable "Gender" to binary
preECMO$Gender <- as.factor(ifelse(preECMO$Gender =="m",0,1))

# extract continuous variables
ndata <- preECMO %>%
  select(-c(Indication1, Indication2, Indication3, Indication4, Indication5, Indication6, Indication7, Gender, ECMO_Survival))

# normalize continuous variables
scaled <- apply(ndata, 2, scale)

# min-max normalization
# maxs <- apply(ndata, 2, max)
# mins <- apply(ndata, 2, min)
# scaled <- as.data.frame(scale(ndata, center = mins, scale = maxs - mins))

# Append binary variables to normalised dataframe
normalised_preECMO <- cbind(scaled, preECMO[,c("Indication1", "Indication2", "Indication3", "Indication4", "Indication5", "Indication6", "Indication7", "Gender", "ECMO_Survival")])

# train-test random splitting (80% of data for training and 20% for testing)
set.seed(123)
index <- sample(1:nrow(normalised_preECMO),round(0.8*nrow(normalised_preECMO)))

# normalised data splitting
train_set <- normalised_preECMO[ index,]
test_set <- normalised_preECMO[-index,]

# Display proportion of outcome variable in train and test set
prop.table(table(train_set$ECMO_Survival))
prop.table(table(test_set$ECMO_Survival))

# extract explanatory variable names
variables_train <- as.character(names(train_set))
variables_train <- variables_train[variables_train != "ECMO_Survival"]

# construct a formula with extracted variable names
formula_train <- as.formula(paste("ECMO_Survival", paste(variables_train, collapse=" + "), sep=" ~ "))
# display formula
formula_train

# original (unnormalised) data splitting: 80% training data and 20% test data
train_set_unnormalised <- preECMO[index,]
test_set_unnormalised <- preECMO[-index,]
# glimpse(train_set_unnormalised)

# Data splitting for methods that require hyperpameter optimisation: 60% training data, 20% validation set and 20% train set
n <-  nrow(normalised_preECMO)
ind1 <- sample(c(1:n),round(n*0.6))
ind2 <- sample(c(1:n)[-ind1],round(n*0.2))
ind3 <- setdiff(c(1:n),c(ind1,ind2))
train.data <- normalised_preECMO[ind1, ]
valid.data <- normalised_preECMO[ind2, ]
test.data  <- normalised_preECMO[ind3, ]
dim(train.data)

# display proportion of outcome variable in each set
prop.table(table(train.data$ECMO_Survival))
prop.table(table(valid.data$ECMO_Survival))
prop.table(table(test.data$ECMO_Survival))

# extract variable names
variables_train1 <- as.character(names(train.data))
variables_train1 <- variables_train1[variables_train1 != "ECMO_Survival"]

# numerical variable in the training set
n_train <- train_set %>%
  select(-c("Indication1", "Indication2", "Indication3", "Indication4", "Indication5", "Indication6", "Indication7", "Gender", "ECMO_Survival"))

# convert type of variable into integers - validation set
n_valid_predictors <- valid.data %>%
  select(-c("ECMO_Survival")) %>%
  mutate(Indication1 = as.integer(Indication1)-1)%>%
  mutate(Indication2 = as.integer(Indication2)-1)%>%
  mutate(Indication3 = as.integer(Indication3)-1)%>%
  mutate(Indication4 = as.integer(Indication4)-1)%>%
  mutate(Indication5 = as.integer(Indication5)-1)%>%
  mutate(Indication6 = as.integer(Indication6)-1)%>%
  mutate(Indication7 = as.integer(Indication7)-1)%>%
  mutate(Gender = as.integer(ifelse(valid.data$Gender=='w',1,0)))

# convert type of variable into integers - test set
n_test_predictors <- test.data %>%
  select(-c("ECMO_Survival")) %>%
  mutate(Indication1 = as.integer(Indication1)-1)%>%
  mutate(Indication2 = as.integer(Indication2)-1)%>%
  mutate(Indication3 = as.integer(Indication3)-1)%>%
  mutate(Indication4 = as.integer(Indication4)-1)%>%
  mutate(Indication5 = as.integer(Indication5)-1)%>%
  mutate(Indication6 = as.integer(Indication6)-1)%>%
  mutate(Indication7 = as.integer(Indication7)-1)%>%
  mutate(Gender = as.integer(ifelse(test.data$Gender=='w',1,0)))



# data splitting for decision trees: these sets contain missing values
train_set_tree <- preECMO_original[ index,]
test_set_tree <- preECMO_original[-index,]


```

# Feature selection

## Lasso regression

```{r vslrgn, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
# extract numerical variables
n_lasso <- normalised_preECMO %>%
  select(-c("Indication1", "Indication2", "Indication3", "Indication4", "Indication5", "Indication6", "Indication7", "Gender", "ECMO_Survival"))
# glimpse(n_lasso)

# construct model matrix for binary variables in train set
xfactors_train_full <- model.matrix(ECMO_Survival ~ Gender + Indication1 + Indication2 + Indication3 + Indication4 + Indication5 + Indication6 + Indication7, data = normalised_preECMO)[, -1]

# predictors matrix (combining numerical and binary variables)
x_full <- as.matrix(data.frame(n_lasso, xfactors_train_full))
# response variable (convert to binary)
y_full <- ifelse(normalised_preECMO$ECMO_Survival == "Survived", 1, 0)


# Fit model: alpha=1 for lasso only and can blend with ridge penalty down to
# alpha=0 ridge only.
set.seed(123)
glmmod01 <- glmnet(x_full, y_full, alpha=1, family="binomial")

# Plot variable coefficients vs. shrinkage parameter lambda.
plot(glmmod01, xvar="lambda")
# glmmod
# coef(glmmod01)[, 10]

# Fit cross-validation model
set.seed(123)
cv.glmmod01 <- cv.glmnet(x_full, y_full, alpha=1, family="binomial")
# Plot results
plot(cv.glmmod01)
# Print minimum value of lambda
best_lambda01 <- cv.glmmod01$lambda.min
# best_lambda01

# Fit model using minimum lambda
best_model01 <- glmnet(x_full, y_full, alpha = 1, lambda = best_lambda01, family="binomial")
# print the value of coefficients
coef(best_model01)

assess.glmnet(best_model01,           #in this case, we are evaluating the model
              newx = x_full,              #in the same data used to fit the model
              newy = y_full, family = "binomial" ) 

# Plot selected variables and their coefficients
coefList <- coef(best_model01, s='lambda.1se')
coefList <- data.frame(coefList@Dimnames[[1]][coefList@i+1],coefList@x) 
names(coefList) <- c('var','val') # construct coefficient list

coefList <- coefList %>%
  arrange(-abs(val)) 
coefList <- coefList[-1,] # remove intercept
# coefList

# plot variable importance
df01 <- data.frame(imp = coefList$val)
rownames(df01) <- coefList$var
df02 <- df01 %>% 
  tibble::rownames_to_column() %>% 
  dplyr::rename("variable" = rowname) %>% 
  dplyr::arrange(imp) %>%
  dplyr::mutate(variable = forcats::fct_inorder(variable))
ggplot(df02) +
  geom_segment(aes(x = variable, y = 0, xend = variable, yend = imp), 
               size = 1.5, alpha = 0.7) +
  geom_point(aes(x = variable, y = imp, col = variable), 
             size = 4, show.legend = F) +
  coord_flip() +
  theme_bw() + labs(title = "Variable selection by LASSO",
              x = "Variable", y = "Coefficient")

# save image to device
# ggsave(
#   "lassoimp.png",
#   plot = last_plot(),
#   device = "png",
#   path = "~/Downloads",
#   scale = 0.5,
#   width = 30,
#   height = 15,
#   units = "cm",
#   dpi = 600,
#   limitsize = TRUE,
#   bg = NULL
# )
```


## Boruta

```{r borutta, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
library(Boruta) # required library

# perform boruta algorrithm on the entire dataset. Maximum number of importance source runs=500
set.seed(111)
boruta <- Boruta(formula_train, data = preECMO, doTrace = 2, maxRuns = 500)
print(boruta) # print results

# attStats(boruta)
# extract names of confirmed important variable
variables_boruta_confirmed01 <- rownames(attStats(boruta)[(which(attStats(boruta)$decision=="Confirmed")),]) 

plot(boruta, las = 2, cex.axis = 0.7) # plot results
plotImpHistory(boruta) # plot history of 500 iterations

# png(file="~/Downloads/boruta1.png",
#   width = 28,
#   height = 13,
#   units = "cm",
#   res = 600)
# plot(boruta, las = 2, cex.axis = 0.7) # plot results
# # a function call to save the file
# dev.off()
# 
# png(file="~/Downloads/boruta2.png",
#   width = 28,
#   height = 13,
#   units = "cm",
#   res = 600)
# plotImpHistory(boruta) # plot history of 500 iterations
# dev.off()

# bor <- TentativeRoughFix(boruta)
# print(bor)

# get formula using variables selected by boruta
rf_formula_boruta <- getNonRejectedFormula(boruta) # formula with non rejected variables
rf_borota_confirmed <- getConfirmedFormula(boruta) # formula with confirmed important variables
```

# Prediction
## Logistic regression

```{r vs1, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
# fit a logistic regression model with all the variables in the dataset
set.seed(123)
glm1 <- glm(formula=ECMO_Survival~., data=train_set_unnormalised, family = binomial)
summary(glm1) # model summary

# generate predictions on test set
glm1_predictions <- predict(glm1, test_set_unnormalised, type="response", family="binomial")
# convert generated prediction to binary 
predicted_classes_glm1 <- ifelse(glm1_predictions > 0.5, "Survived", "Died")

# create confusion matrix
cmatrix_glm1 <- confusionMatrix(data= as.factor(predicted_classes_glm1), reference = as.factor(test_set_unnormalised$ECMO_Survival), dnn = c("Prediction", "Original"))
cmatrix_glm1
#Display results
cmatrix_glm1$byClass

mcc_glm1 <- mcc(as.factor(test_set_unnormalised$ECMO_Survival), as.factor(predicted_classes_glm1))
mcc_glm1 # compute MCC
accuracy_glm1 <- cmatrix_glm1$overall['Accuracy']   
accuracy_glm1 # Accuracy of the model
f1_glm1 <- cmatrix_glm1$byClass['F1']   
f1_glm1 # F1 score of the model

# extract variables selected by Lasso regression
toselect.x0 <- coefList$var
toselect.x0 = gsub(pattern = "1", replacement = "", x = toselect.x0) # rename

# change entry type to character
relevant.x0 <- as.character((toselect.x0))
# relevant.x0

# construct formula with only sig variables
sig.formula0 <- as.formula(paste("ECMO_Survival", paste(relevant.x0, collapse=" + "), sep=" ~ "))

# fit logistic regression model with variable selected by Lasso
set.seed(123)
sig.model0 <- glm(formula=sig.formula0, data=train_set_unnormalised, family = binomial)
summary(sig.model0)

# generate predictions
glm_predictions <- predict(sig.model0, test_set_unnormalised, type="response", family="binomial")
# convert generated prediction to binary
predicted_classes_glm <- ifelse(glm_predictions > 0.5, "Survived", "Died")

# create confusion matrix
cmatrix_glm <- confusionMatrix(data= as.factor(predicted_classes_glm), reference = as.factor(test_set_unnormalised$ECMO_Survival),
                           dnn = c("Prediction", "Original"))
cmatrix_glm
#Display results
cmatrix_glm$byClass

mcc_glm <- mcc(as.factor(test_set_unnormalised$ECMO_Survival), as.factor(predicted_classes_glm))
mcc_glm # Calculate MCC
accuracy_glm <- cmatrix_glm$overall['Accuracy']   
accuracy_glm # Accuracy of the model
f1_glm <- cmatrix_glm$byClass['F1']   
f1_glm # F1 score of the model
```

## Decision trees

```{r dtree, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
# load required libraries
library(rpart)
library(rpart.plot)

# build first decision tree
set.seed(123)
first_tree <- rpart(formula_train, data=train_set_tree, method="class")
rpart.plot(first_tree,type=2, extra=106, main="First tree") # plot the tree

# create a plot showing variable importance
df01 <- data.frame(imp = first_tree$variable.importance)
df02 <- df01 %>% 
  tibble::rownames_to_column() %>% 
  dplyr::rename("variable" = rowname) %>% 
  dplyr::arrange(imp) %>%
  dplyr::mutate(variable = forcats::fct_inorder(variable))
ggplot(df02) +
  geom_segment(aes(x = variable, y = 0, xend = variable, yend = imp), 
               size = 1.5, alpha = 0.7) +
  geom_point(aes(x = variable, y = imp, col = variable), 
             size = 4, show.legend = F) +
  coord_flip() +
  theme_bw()

printcp(first_tree) # print cp of the first tree

cp.select <- function(big.tree) {
  min.x <- which.min(big.tree$cptable[, 4]) #column 4 is xerror
  big.tree$cptable[min.x, 1]
}

# prune first tree
first_tree_pruned <- prune(first_tree, cp=cp.select(first_tree)+0.001)
rpart.plot(first_tree_pruned, extra=106, main="Pruned tree") # plot the tree

cp.select(first_tree)

# function for accuracy tune
accuracy_tune <- function(fit) {
    predict_unseen <- predict(fit, test_set_tree, type = 'class')
    table_mat <- table(test_set_tree$ECMO_Survival, predict_unseen)
    accuracy_Test <- sum(diag(table_mat)) / sum(table_mat)
    return(accuracy_Test)
}

paste("Accuracy of first tree: ", accuracy_tune(first_tree))
paste("Accuracy of pruned tree: ", accuracy_tune(first_tree_pruned))

# settings for tuning the tree
control <- rpart.control(minsplit = 4,
    minbucket = round(5 / 3),
    maxdepth = 3,
    cp = cp.select(first_tree)+0.001)

# build a tree using control values
tune_fit <- rpart(formula_train, data = train_set_tree, method = 'class', control = control)
print(paste('Accuracy for tuned fit: ',accuracy_tune(tune_fit)))

# generate predictions
tree_prediction <- predict(first_tree_pruned, newdata=test_set_tree, type="class")

# calculate classification rate
overall_class_rate_tree <- mean(tree_prediction==test_set_tree$ECMOSurvival)

#Create confusion matrix
cmatrix_tree <- confusionMatrix(data=as.factor(tree_prediction), reference = as.factor(test_set_tree$ECMO_Survival),  dnn = c("Prediction", "Original"))
cmatrix_tree
#Display results 
cmatrix_tree$byClass

mcc_tree_full <- mcc(as.factor(test_set_tree$ECMO_Survival), as.factor(tree_prediction))
mcc_tree_full # calculate MCC
library(MLmetrics)
f1_tree_full <- F1_Score(as.factor(test_set_tree$ECMO_Survival), as.factor(tree_prediction))
f1_tree_full # F1 score
accuracy_tree_full <- cmatrix_tree$overall['Accuracy']   
accuracy_tree_full # accuracy

# build decision tree using variables selected by boruta
set.seed(123)
boruta_dtree <- rpart(rf_borota_confirmed, data=train_set_tree, method="class")
rpart.plot(first_tree,type=2,extra=106, main="Boruta tree") # plot the tree
paste("Accuracy of boruta tree: ",accuracy_tune(boruta_dtree)) # accuracy

# generate predictions
tree_predictionB <- predict(boruta_dtree, newdata=test_set_tree, type="class")
# calculate classification rate
overall_class_rate_treeB <- mean(tree_predictionB==test_set_tree$ECMOSurvival)

#Create confusion matrix
cmatrix_treeB <- confusionMatrix(data=as.factor(tree_predictionB), reference = as.factor(test_set_tree$ECMO_Survival),
                           dnn = c("Prediction", "Original"))
cmatrix_treeB
#Display results 
cmatrix_treeB$byClass

mcc_btree <- mcc(as.factor(test_set_tree$ECMO_Survival), as.factor(tree_predictionB))
mcc_btree # calculate MCC
accuracy_btree <- cmatrix_treeB$overall['Accuracy']   
accuracy_btree # accuracy
f1_btree <- cmatrix_treeB$byClass['F1']   
f1_btree # f1 score

```

## Random forest

```{r rforest, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
# load required libraries
library(randomForest)
library(forcats)

# build a random forest using all the variables in the dataset
set.seed(123)
rf_train <-  randomForest(formula_train, data=train_set, ntree = 500) # no. of trees=500
rf_train

# generate predictions 
predictions_randomF <- predict(rf_train, test_set, "class")

#Create confusion matrix
cmatrix_rf <- confusionMatrix(data=as.factor(predictions_randomF), reference = as.factor(test_set$ECMO_Survival),dnn = c("Prediction", "Original"))
cmatrix_rf
cmatrix_rf$byClass

mcc_rf <- mcc(as.factor(test_set$ECMO_Survival), as.factor(predictions_randomF))
mcc_rf # calculate MCC
accuracy_rf <- cmatrix_rf$overall['Accuracy']   
accuracy_rf # accuracy
f1_rf <- cmatrix_rf$byClass['F1']   
f1_rf # f1 score

# build a random forest using variables selected by boruta
set.seed(123)
rfboruta_c <-  randomForest(rf_borota_confirmed, data=train_set, ntree = 500)
rfboruta_c

# generate predictions
predictions_randomF_borutaC <- predict(rfboruta_c, test_set, "class")

# create confusion matrix
cmatrix_rfc <- confusionMatrix(predictions_randomF_borutaC, test_set$ECMO_Survival)
cmatrix_rfc
cmatrix_rfc$byClass

mcc_rfb <- mcc(as.factor(test_set$ECMO_Survival), as.factor(predictions_randomF_borutaC))
mcc_rfb # calculate MCC
accuracy_rfb <- cmatrix_rfc$overall['Accuracy']   
accuracy_rfb # accuracy
f1_rfb <- cmatrix_rfc$byClass['F1']   
f1_rfb # f1 score
```

## Neural network

```{r nn1, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
# Load required packages
library(neuralnet)
library(NeuralNetTools)
library(nnet)
library(tibble)

# construct a formula for implementing neural networks
formula_nn <- as.formula(paste(" ", paste(variables_train1, collapse=" + "), sep=" ~ "))
# create model matrix
nn_matrix <- model.matrix(formula_nn, data=train.data)[,-1]
# head(nn_matrix,4)              
# Construct predictor list
predictor_list <- paste(colnames(nn_matrix),collapse="+")

# symbolic description of the model we want to fit
f <- paste(c("train.data$ECMO_Survival~",predictor_list),collapse="")


# convert the variable ECMO_Survival into integer
train.data$ECMO_Survival <- as.integer(train.data$ECMO_Survival)-1
valid.data$ECMO_Survival <- as.integer(valid.data$ECMO_Survival)-1
test.data$ECMO_Survival <- as.integer(test.data$ECMO_Survival)-1
# glimpse(train.data)

# train neural networks, generate prediction results using validation set and calculate MCC of each model. learning rate of all the networks set to 0.01.
set.seed(123)
nn_first <- neuralnet(f,data=nn_matrix, hidden=c(5), learningrate=0.01,
linear.output=FALSE, err.fct='ce', likelihood=TRUE) # one hidden layer with 5 nodes
pr.nn1 <- compute(nn_first, n_valid_predictors) # generate prediction on validation set
predictions_test1 <- ifelse(pr.nn1$net.result<0.5,0,1)
mcc_nn1 <- mcc(as.factor(valid.data$ECMO_Survival), as.factor(predictions_test1)) # calculate MCC

set.seed(123)
nn_second <- neuralnet(f,data=nn_matrix, hidden=c(10), learningrate=0.01,
linear.output=FALSE, err.fct='ce', likelihood=TRUE) # one hidden layer with 10 nodes
pr.nn2 <- compute(nn_second, n_valid_predictors) # generate prediction on validation set
predictions_test2 <- ifelse(pr.nn2$net.result<0.5,0,1)
mcc_nn2 <- mcc(as.factor(valid.data$ECMO_Survival), as.factor(predictions_test2)) # calculate MCC

set.seed(123)
nn_two_layers_1 <- neuralnet(f,data=nn_matrix, hidden=c(10,5), learningrate=0.01,
linear.output=FALSE, err.fct='ce', likelihood=TRUE) # two hidden layer with 10 nodes in first layer and 5 nodes in the second
pr.nn3 <- compute(nn_two_layers_1, n_valid_predictors) # generate prediction on validation set
predictions_test3 <- ifelse(pr.nn3$net.result<0.5,0,1)
mcc_nn3 <- mcc(as.factor(valid.data$ECMO_Survival), as.factor(predictions_test3)) # calculate MCC

set.seed(123)
nn_two_layers_2 <- neuralnet(f,data=nn_matrix, hidden=c(10,75), learningrate=0.01,
linear.output=FALSE, err.fct='ce', likelihood=TRUE) # two hidden layer with 10 nodes in first layer and 75 nodes in the second
pr.nn4 <- compute(nn_two_layers_2, n_valid_predictors) # generate prediction on validation set
predictions_test4 <- ifelse(pr.nn4$net.result<0.5,0,1)
mcc_nn4 <- mcc(as.factor(valid.data$ECMO_Survival), as.factor(predictions_test4)) # calculate MCC

# create a dataframe with the name of neural network and their coresponding MCC value
nn_df <- data.frame (network  = c("nn_first", "nn_second", "nn_two_layers_1", "nn_two_layers_2"),
                  mcc = c(mcc_nn1, mcc_nn2, mcc_nn3, mcc_nn4))
nn_df

# function to selectt the model with highest MCC value
nn.select <- function(nn_level) {
  min.x <- which.max(nn_level$mcc) #column 4 is xerror
  nn_level[min.x, 1]
}

nn_chosen <- get(nn.select(nn_df)) # choose the model with highest MCC value
paste("Chosen model:", nn.select(nn_df))

# generate predictions using the selected model applied to test set
pr.nn <- compute(nn_chosen, n_test_predictors)
predictions_test <- ifelse(pr.nn$net.result<0.5,0,1)

#Create confusion matrix
cmatrix_nn <- confusionMatrix(data=as.factor(predictions_test), reference = as.factor(test.data$ECMO_Survival),dnn = c("Prediction", "Original"))
cmatrix_nn
cmatrix_nn$byClass

mcc_nn <- mcc(as.factor(test.data$ECMO_Survival), as.factor(predictions_test))
mcc_nn # calculate MCC
accuracy_nn <- cmatrix_nn$overall['Accuracy']   
accuracy_nn # accuracy
f1_nn <- cmatrix_nn$byClass['F1']   
f1_nn # F1 score

```

## Naive Bayes

```{r nb, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
library(e1071) # load required packages

# build a naive Bayes classifier using all the variables in the dataset
set.seed(123)  # Setting Seed
classifier_nb <- naiveBayes(formula_train, data = train_set)
classifier_nb

# Predicting on test data'
y_pred <- predict(classifier_nb, newdata = test_set, type="class")

# Confusion Matrix
cm <- table(test_set$ECMO_Survival, y_pred)
cm
 
# Model Evaluation
cmatrix_nb <- confusionMatrix(cm)
cmatrix_nb
cmatrix_nb$byClass

mcc_nb <- mcc(as.factor(test_set$ECMO_Survival), as.factor(y_pred))
mcc_nb # calculate MCC
accuracy_nb <- cmatrix_nb$overall['Accuracy']   
accuracy_nb # accuracy
f1_nb <- cmatrix_nb$byClass['F1']   
f1_nb # f1 score

# build a second naive Bayes using only the variables selected by Lasso
set.seed(123)  # Setting Seed
classifier_nb2 <- naiveBayes(sig.formula0, data = train_set)
classifier_nb2

# Predicting on test data'
y_pred2 <- predict(classifier_nb2, newdata = test_set, type="class")

# Confusion Matrix
cm2 <- table(test_set$ECMO_Survival, y_pred)
cm2
 
# Model Evaluation
cmatrix_nb2 <- confusionMatrix(cm2)
cmatrix_nb2
cmatrix_nb2$byClass

mcc_nb2 <- mcc(as.factor(test_set$ECMO_Survival), as.factor(y_pred2))
mcc_nb2 # calculate MCC
accuracy_nb2 <- cmatrix_nb2$overall['Accuracy']   
accuracy_nb2 # accuracy
f1_nb2 <- cmatrix_nb2$byClass['F1']   
f1_nb2 # f1 score

```

## Gradient boosting

```{r boost, echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
library(xgboost) # load required packages

# convert response variable to binary
train_set_unnormalised$ECMO_Survival <- ifelse(train_set_unnormalised$ECMO_Survival=="Survived",1,0)
test_set_unnormalised$ECMO_Survival <- ifelse(test_set_unnormalised$ECMO_Survival=="Survived",1,0)

# prepare training data (convert to a data matrix)
train_data_xg <- data.matrix(train_set_unnormalised %>% select(-c("ECMO_Survival")))
train_labels_xg <- train_set_unnormalised$ECMO_Survival # train data labels
dtrain <- xgb.DMatrix(data = train_data_xg, label= train_labels_xg)

# prepare testing data (convert to a data matrix)
test_data_xg <- data.matrix(test_set_unnormalised %>% select(-c("ECMO_Survival")))
test_labels_xg <- test_set_unnormalised$ECMO_Survival # test data labels
dtest <- xgb.DMatrix(data = test_data_xg, label= test_labels_xg)

# define negative case and positive case
negative_cases <- sum(train_labels_xg == 1)
postive_cases <- sum(train_labels_xg == 0)

# fit gradient boosted model
set.seed(123)
modelxg <- xgboost(data = dtrain, # the data  
                 nround = 2, # max number of boosting iterations
                 max.depth = 4, # the maximum depth of each decision tree
                 objective = "binary:logistic",  # the objective function
                 scale_pos_weight = negative_cases/postive_cases) # control for imbalanced classes

# get information on how important each feature is
importance_matrixxg <- xgb.importance(names(train_set_unnormalised %>% select(-c("ECMO_Survival"))), model = modelxg)

# plot it
xgb.plot.importance(importance_matrixxg)

# generate predictions
predxg <- predict(modelxg, dtest)
# get & print the classification error
err <- mean(as.numeric(predxg > 0.5) != test_labels_xg)
print(paste("test-error=", err))

# Create confusion Matrix
cmxg <- table(as.factor(test_labels_xg), as.factor(ifelse(predxg<0.5,0,1)))
cmxg
 
# Model Evaluation
cmatrix_xg <- confusionMatrix(cmxg)
cmatrix_xg
cmatrix_xg$byClass

mcc_xg <- mcc(as.factor(test_labels_xg), as.factor(ifelse(predxg<0.5,0,1)))
mcc_xg # calculate MCC
accuracy_xg <- cmatrix_xg$overall['Accuracy']   
accuracy_xg # accuracy
f1_xg <- cmatrix_xg$byClass['F1']   
f1_xg # f1 score

```

```{r metricPlot, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
### plot MCC, F1 score and accuracy of all the above models

# create a dataframe with variable names and their corresponding values of metrics
metric_all <- tibble(Model = rep(c("GLM", "GLM lasso", "DT", "DT boruta", "RF", "RF boruta", "NN", "NB", "NB lasso", "XGboost"), each = 3),
Metric = rep(c("MCC", "F1", "Accuracy"), time = 10),
Value = c(mcc_glm1, f1_glm1, accuracy_glm1, mcc_glm, f1_glm, accuracy_glm, mcc_tree_full, f1_tree_full, accuracy_tree_full, mcc_btree, f1_btree, accuracy_btree,
        mcc_rf, f1_rf, accuracy_rf, mcc_rfb, f1_rfb, accuracy_rfb,
        mcc_nn, f1_nn, accuracy_nn, mcc_nb, f1_nb, accuracy_nb,mcc_nb2, f1_nb2, accuracy_nb2,
        mcc_xg, f1_xg, accuracy_xg))
metric_all

# plot the data
metric_plot <- metric_all %>%
ggplot(aes(Model, Value, fill = Metric)) +
geom_col(position = "dodge")
metric_plot


# ggsave(
#   "metricplot1.png",
#   plot = last_plot(),
#   device = "png",
#   path = "~/Downloads",
#   scale = 0.5,
#   width = 40,
#   height = 15,
#   units = "cm",
#   dpi = 600,
#   limitsize = TRUE,
#   bg = NULL
# )
```

# Under sampling

```{r underSampling, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
library(ROSE) # load required packages
table(preECMO$ECMO_Survival) # display proportion of outcome variable in imbalanced dataset

# sample data: new balanced dataset contain all the instances in minority class 
sampled_data <- ovun.sample(ECMO_Survival ~ ., data = preECMO, method = "under", p=0.5, seed = 123)$data

table(sampled_data$ECMO_Survival) # display proportion of outcome variable in undersampled dataset

# train-test random splitting (80% of data for training and 20% for testing)
set.seed(123)
index2 <- sample(1:nrow(sampled_data),round(0.8*nrow(sampled_data)))

# normalised data splitting
train_set_sampled <- sampled_data[ index2,]
test_set_sampled <- sampled_data[-index2,]

glimpse(train_set_sampled)
# check proportion of survival indicator in train and test set
prop.table(table(train_set_sampled$ECMO_Survival))
prop.table(table(test_set_sampled$ECMO_Survival))

ndata_sampled <- sampled_data %>%
  select(-c(Indication1, Indication2, Indication3, Indication4, Indication5, Indication6, Indication7, Gender, ECMO_Survival))

# normalize continuous variables
scaled_sampled <- apply(ndata_sampled, 2, scale)

# normalised dataframe
normalised_sampled <- cbind(scaled_sampled, sampled_data[,c("Indication1", "Indication2", "Indication3", "Indication4", "Indication5", "Indication6", "Indication7", "Gender", "ECMO_Survival")])

# normalised data splitting
train_set_sampled_normalised <- normalised_sampled[ index2,]
test_set_sampled_normalised <- normalised_sampled[-index2,]

# data splitting for neural networks (60% train , 20 % validation and 20% test)
n_s <-  nrow(normalised_sampled)
ind1s <- sample(c(1:n_s),round(n_s*0.6))
ind2s <- sample(c(1:n_s)[-ind1],round(n_s*0.2))
ind3s <- setdiff(c(1:n_s),c(ind1,ind2))
train.data_sampled <- normalised_sampled[ind1s, ]
valid.data_sampled <- normalised_sampled[ind2s, ]
test.data_sampled  <- normalised_sampled[ind3s, ]

# proportion of response variable in train, valid and test set
prop.table(table(train.data_sampled$ECMO_Survival))
prop.table(table(valid.data_sampled$ECMO_Survival))
prop.table(table(test.data_sampled$ECMO_Survival))

# convert type of variable into integers - train set
train.data_sampled <- train.data_sampled %>%
  mutate(Indication1 = as.integer(Indication1)-1)%>%
  mutate(Indication2 = as.integer(Indication2)-1)%>%
  mutate(Indication3 = as.integer(Indication3)-1)%>%
  mutate(Indication4 = as.integer(Indication4)-1)%>%
  mutate(Indication5 = as.integer(Indication5)-1)%>%
  mutate(Indication6 = as.integer(Indication6)-1)%>%
  mutate(Indication7 = as.integer(Indication7)-1)%>%
  mutate(Gender = as.integer(ifelse(train.data_sampled$Gender=='w',1,0)))

# convert type of variable into integers - validation set predictors
n_valid_predictors_sampled <- valid.data_sampled %>%
  select(-c("ECMO_Survival")) %>%
  mutate(Indication1 = as.integer(Indication1)-1)%>%
  mutate(Indication2 = as.integer(Indication2)-1)%>%
  mutate(Indication3 = as.integer(Indication3)-1)%>%
  mutate(Indication4 = as.integer(Indication4)-1)%>%
  mutate(Indication5 = as.integer(Indication5)-1)%>%
  mutate(Indication6 = as.integer(Indication6)-1)%>%
  mutate(Indication7 = as.integer(Indication7)-1)%>%
  mutate(Gender = as.integer(ifelse(valid.data_sampled$Gender=='w',1,0)))

# convert type of variable into integers - test set predictors
n_test_predictors_sampled <- test.data_sampled %>%
  select(-c("ECMO_Survival")) %>%
  mutate(Indication1 = as.integer(Indication1)-1)%>%
  mutate(Indication2 = as.integer(Indication2)-1)%>%
  mutate(Indication3 = as.integer(Indication3)-1)%>%
  mutate(Indication4 = as.integer(Indication4)-1)%>%
  mutate(Indication5 = as.integer(Indication5)-1)%>%
  mutate(Indication6 = as.integer(Indication6)-1)%>%
  mutate(Indication7 = as.integer(Indication7)-1)%>%
  mutate(Gender = as.integer(ifelse(test.data_sampled$Gender=='w',1,0)))
```



## Prediction

## Logistic regression

```{r glms, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
# convert response variable in the train and test set to binary
train_set_sampled$ECMO_Survival <- ifelse(train_set_sampled$ECMO_Survival=="Survived",1,0)
test_set_sampled$ECMO_Survival <- ifelse(test_set_sampled$ECMO_Survival=="Survived",1,0)

# View dataset
# glimpse(train_set_sampled)

# fit model
set.seed(123)
glm_us <- glm(formula_train, data=train_set_sampled, family = binomial)
summary(glm_us) # print summary

# Generate predictions
glm_us_predictions <- predict(glm_us, test_set_sampled, type="response", family="binomial")

# Convert generated probability to binary response
predicted_classes_glm_us <- ifelse(glm_us_predictions > 0.5, 1, 0)

# create confusion matrix
cmatrix_glm_us <- confusionMatrix(data= as.factor(predicted_classes_glm_us), reference = as.factor(test_set_sampled$ECMO_Survival), dnn = c("Prediction", "Original"))
cmatrix_glm_us
#Display results
cmatrix_glm_us$byClass

mcc_glm_us <- mcc(as.factor(test_set_sampled$ECMO_Survival), as.factor(predicted_classes_glm_us))
mcc_glm_us # calculate MCC
accuracy_glm_us <- cmatrix_glm_us$overall['Accuracy']   
accuracy_glm_us # accuracy
f1_glm_us <- cmatrix_glm_us$byClass['F1']   
f1_glm_us # f1 score


# fit model using lasso-selected variables
set.seed(123)
glm_us2 <- glm(formula=sig.formula0, data=train_set_sampled, family = binomial)
summary(glm_us2) # print model summary

# Generate predictions
glm_us2_predictions <- predict(glm_us2, test_set_sampled, type="response", family="binomial")
# Convert generated probability to binary response
predicted_classes_glm_us2 <- ifelse(glm_us2_predictions > 0.5, 1, 0)

# create confusion matrix
cmatrix_glm_us2 <- confusionMatrix(data= as.factor(predicted_classes_glm_us2), reference = as.factor(test_set_sampled$ECMO_Survival),
                           dnn = c("Prediction", "Original"))
cmatrix_glm_us2
#Display results
cmatrix_glm_us2$byClass

mcc_glm_us2 <- mcc(as.factor(test_set_sampled$ECMO_Survival), as.factor(predicted_classes_glm_us2))
mcc_glm_us2 # calculate MCC
accuracy_glm_us2 <- cmatrix_glm_us2$overall['Accuracy']   
accuracy_glm_us2 # accuracy
f1_glm_us2 <- cmatrix_glm_us2$byClass['F1']   
f1_glm_us2 # f1 score


```

### Decision trees

```{r dtreeU, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
# Rename response variable levels in both train and test set for clarity
train_set_sampled$ECMO_Survival <- as.factor(ifelse(train_set_sampled$ECMO_Survival==1, "Survived","Died"))
test_set_sampled$ECMO_Survival <- as.factor(ifelse(test_set_sampled$ECMO_Survival==1, "Survived", "Died"))

# model fitting: first tree
set.seed(123)
first_tree_sampled <- rpart(formula_train, data=train_set_sampled, method="class")
rpart.plot(first_tree_sampled,type=2,extra=106) # plot tree

printcp(first_tree_sampled) # print cp
# select cp with mininum xerror
cp.select(first_tree_sampled)

# set control
control <- rpart.control(minsplit = 4,
    minbucket = round(7 / 3),
    maxdepth = 3)

#tune_fit_sampled <- rpart(formula_train, data = train_set_sampled, method = 'class', cp=cp.select(first_tree_sampled))

# prune tree
tune_fit_sampled <- prune(first_tree_sampled, cp=cp.select(first_tree_sampled))
# generate predictions
tree_prediction_sampled <- predict(first_tree_sampled, newdata=test_set_sampled, type="class")

#Create confusion matrix
cmatrix_tree_sampled <- confusionMatrix(data=as.factor(tree_prediction_sampled), reference = as.factor(test_set_sampled$ECMO_Survival),dnn = c("Prediction", "Original"))

cmatrix_tree_sampled
cmatrix_tree_sampled$byClass

mcc_dts <- mcc(as.factor(test_set_sampled$ECMO_Survival), as.factor(tree_prediction_sampled))
mcc_dts # calculate MCC
accuracy_dts <- cmatrix_tree_sampled$overall['Accuracy']   
accuracy_dts # accuracy
f1_dts <- cmatrix_tree_sampled$byClass['F1']   
f1_dts # f1 score

# model fitting: using Boruta-selected variables
set.seed(123)
boruta_dtree_sampled <- rpart(rf_borota_confirmed, data=train_set_sampled, method="class")
rpart.plot(boruta_dtree_sampled,type=2,extra=106) # plot tree
printcp(boruta_dtree_sampled) # print cp

# generate predictions
tree_prediction_boruta_sampled <- predict(boruta_dtree_sampled, newdata=test_set_sampled, type="class")

#Create confusion matrix
cmatrix_tree_boruta_sampled <- confusionMatrix(data=as.factor(tree_prediction_boruta_sampled), reference = as.factor(test_set_sampled$ECMO_Survival),
                           dnn = c("Prediction", "Original"))
cmatrix_tree_boruta_sampled
#Display results 
cmatrix_tree_boruta_sampled$byClass

mcc_bdts <- mcc(as.factor(test_set_sampled$ECMO_Survival), as.factor(tree_prediction_boruta_sampled))
mcc_bdts # calculate MCC
accuracy_bdts <- cmatrix_tree_boruta_sampled$overall['Accuracy']   
accuracy_bdts # accuracy
f1_bdts <- cmatrix_tree_boruta_sampled$byClass['F1']   
f1_bdts #f1 score

```

### Random forest

```{r rforestunder, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
# fit random forest using all the variables in the dataset
set.seed(123)
rf_train_sampled <-  randomForest(formula_train, data=train_set_sampled, ntree = 500)
rf_train_sampled
# generate predictions
predictions_randomF_sampled <- predict(rf_train_sampled, test_set_sampled, "class")
predictions_randomF_sampled

#Create confusion matrix
cmatrix_rf_sampled <- confusionMatrix(data=as.factor(predictions_randomF_sampled), reference = as.factor(test_set_sampled$ECMO_Survival), dnn = c("Prediction", "Original"))

cmatrix_rf_sampled
cmatrix_rf_sampled$byClass

mcc_rfs <- mcc(as.factor(test_set_sampled$ECMO_Survival), as.factor(predictions_randomF_sampled))
mcc_rfs # calculate MCC
accuracy_rfs <- cmatrix_rf_sampled$overall['Accuracy']   
accuracy_rfs # accuracy
f1_rfs <- cmatrix_rf_sampled$byClass['F1']   
f1_rfs # f1 score

# fit random forest using Boruta-selected variables
set.seed(123)
rf_confirmed_boruta_sampled <-  randomForest(rf_borota_confirmed, data=train_set_sampled, ntree = 500)
rf_confirmed_boruta_sampled

# Generate predictions
predictions_randomF_borutaC_sampled <- predict(rf_confirmed_boruta_sampled, test_set_sampled, "class")

# Create confusion matrix
cmatrix_rf_cboruta_sampled <- confusionMatrix(predictions_randomF_borutaC_sampled, test_set_sampled$ECMO_Survival)

cmatrix_rf_cboruta_sampled
cmatrix_rf_cboruta_sampled$byClass

# calculate MCC
mcc_brfs <- mcc(as.factor(test_set_sampled$ECMO_Survival), as.factor(predictions_randomF_borutaC_sampled))
mcc_brfs
accuracy_brfs <- cmatrix_rf_cboruta_sampled$overall['Accuracy']   
accuracy_brfs # accuracy
f1_brfs <- cmatrix_rf_cboruta_sampled$byClass['F1']   
f1_brfs # f1 score


```

## Neural network

```{r nns, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
# Create model matrix 
nn_matrix_sampled <- model.matrix(formula_nn, data=train.data_sampled)[,-1]
# head(nn_matrix,4)     

# Construct predictor list
predictor_list_sampled <- paste(colnames(nn_matrix_sampled),collapse="+")

# symbolic description of the model we want to fit
fs <- paste(c("train.data_sampled$ECMO_Survival~",predictor_list_sampled),collapse="")


# convert the variable successful into integer
train.data_sampled$ECMO_Survival <- ifelse(train.data_sampled$ECMO_Survival == "Survived",1,0)
valid.data_sampled$ECMO_Survival <- ifelse(valid.data_sampled$ECMO_Survival == 'Survived',1,0)
test.data_sampled$ECMO_Survival <- ifelse(test.data_sampled$ECMO_Survival == 'Survived',1,0)
glimpse(train.data_sampled)

# fit model
set.seed(123)
nn_first_s <- neuralnet(fs,data=nn_matrix_sampled, hidden=c(5), learningrate=0.01,
linear.output=FALSE, err.fct='ce', likelihood=TRUE) # one hidden layer five nodes
pr.nn1s <- compute(nn_first_s, n_valid_predictors_sampled) # generate predictions in validation set
predictions_test1s <- ifelse(pr.nn1s$net.result<0.5,0,1)
mcc_nn1s <- mcc(as.factor(valid.data_sampled$ECMO_Survival), as.factor(predictions_test1s)) # calculate MCC

set.seed(123)
nn_seconds <- neuralnet(fs,data=nn_matrix_sampled, hidden=c(10), learningrate=0.01,
linear.output=FALSE, err.fct='ce', likelihood=TRUE) # one hidden layer 10 nodes
pr.nn2s <- compute(nn_seconds, n_valid_predictors_sampled) # generate predictions in validation set
predictions_test2s <- ifelse(pr.nn2s$net.result<0.5,0,1)
mcc_nn2s <- mcc(as.factor(valid.data_sampled$ECMO_Survival), as.factor(predictions_test2s)) # calculate MCC

set.seed(123)
nn_two_layers_1s <- neuralnet(fs,data=nn_matrix_sampled, hidden=c(10,5), learningrate=0.01,
linear.output=FALSE, err.fct='ce', likelihood=TRUE) # two hidden layers with 10 nodes in first layer and five nodes in second
pr.nn3s <- compute(nn_two_layers_1s, n_valid_predictors_sampled) # generate predictions in validation set
predictions_test3s <- ifelse(pr.nn3s$net.result<0.5,0,1) 
mcc_nn3s <- mcc(as.factor(valid.data_sampled$ECMO_Survival), as.factor(predictions_test3s)) # calculate MCC

set.seed(123)
nn_two_layers_2s <- neuralnet(fs,data=nn_matrix_sampled, hidden=c(10,75), learningrate=0.01,
linear.output=FALSE, err.fct='ce', likelihood=TRUE) # two hidden layers with 10 nodes in first layer and 75 nodes in second
pr.nn4s <- compute(nn_two_layers_2s, n_valid_predictors_sampled) # generate predictions in validation set
predictions_test4s <- ifelse(pr.nn4s$net.result<0.5,0,1)
mcc_nn4s <- mcc(as.factor(valid.data_sampled$ECMO_Survival), as.factor(predictions_test4s)) # calculate MCC

# Create a dataframe with neural networks and thir corresponding MCC values
nn_dfs <- data.frame (network  = c("nn_first", "nn_second", "nn_two_layers_1", "nn_two_layers_2"),
                  mcc = c(mcc_nn1s, mcc_nn2s, mcc_nn3s, mcc_nn4s))
nn_dfs

# choose model with highest MCC
paste("Chosen model:", nn.select(nn_dfs))
nn_chosens <- get(nn.select(nn_dfs))

# generate predictions using test set
pr.nns <- compute(nn_chosens, n_test_predictors_sampled)
predictions_test_sampled <- ifelse(pr.nns$net.result<0.5,0,1) # convert prrobability to binary response

#Create confusion matrix
cmatrix_nn_sampled <- confusionMatrix(data=as.factor(predictions_test_sampled), reference = as.factor(test.data_sampled$ECMO_Survival),dnn = c("Prediction", "Original"))
cmatrix_nn_sampled
cmatrix_nn_sampled$byClass

mcc_nns <-mcc(as.factor(test.data_sampled$ECMO_Survival), as.factor(predictions_test_sampled))
mcc_nns # calculate MCC
accuracy_nns <- cmatrix_nn_sampled$overall['Accuracy']   
accuracy_nns # accuracy
f1_nns <- cmatrix_nn_sampled$byClass['F1']   
f1_nns # f1 score

```

## Naive Bayes

```{r nbs, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
# fit model using all the variables in dataset
set.seed(123)  # Setting Seed
classifier_cls <- naiveBayes(formula_train, data = train_set_sampled_normalised)
classifier_cls

# Predicting on test data
y_preds <- predict(classifier_cls, newdata = test_set_sampled_normalised, type="class")

# Create confusion Matrix
cms <- table(test_set_sampled_normalised$ECMO_Survival, y_preds)
cms
 
# Model Evaluation
cmatrix_nbs <- confusionMatrix(cms)
cmatrix_nbs
cmatrix_nbs$byClass
mcc_nbs <- mcc(as.factor(test_set_sampled_normalised$ECMO_Survival), as.factor(y_preds))
mcc_nbs # calculate MCC
accuracy_nbs <- cmatrix_nbs$overall['Accuracy']   
accuracy_nbs # accuracy
f1_nbs <- cmatrix_nbs$byClass['F1']   
f1_nbs # f1 score

# Fit model using variables selected by Lasso
set.seed(123)  # Setting Seed
classifier_cls2 <- naiveBayes(sig.formula0, data = train_set_sampled_normalised)
classifier_cls2

# Predicting on test data
y_preds2 <- predict(classifier_cls2, newdata = test_set_sampled_normalised, type="class")

# Confusion Matrix
cms2 <- table(test_set_sampled_normalised$ECMO_Survival, y_preds2)
cms2
 
# Model Evaluation
cmatrix_nbs2 <- confusionMatrix(cms2)
cmatrix_nbs2
cmatrix_nbs2$byClass
mcc_nbs2 <- mcc(as.factor(test_set_sampled_normalised$ECMO_Survival), as.factor(y_preds2))
mcc_nbs2 # calculate MCC
accuracy_nbs2 <- cmatrix_nbs2$overall['Accuracy']   
accuracy_nbs2 # accuracy
f1_nbs2 <- cmatrix_nbs2$byClass['F1']   
f1_nbs2 # f1 score

```

```{r metricPlots, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
# Plot evaluation metrics values of all the models applied to under-sampled dataset

# Create a dataframe with technique and their metrics
metric_all_sampled <- tibble(Model = rep(c("GLM", "GLM lasso", "DT", "DT boruta", "RF", "RF boruta", "NN", "NB", "NB lasso"), each = 3),
Metric = rep(c("MCC", "F1", "Accuracy"), time = 9),
Value = c(mcc_glm_us, f1_glm_us, accuracy_glm_us, mcc_glm_us2, f1_glm_us2, accuracy_glm_us2,
          mcc_dts, f1_dts, accuracy_dts, mcc_bdts, f1_bdts, accuracy_bdts,
          mcc_rfs, f1_rfs, accuracy_rfs, mcc_brfs, f1_brfs, accuracy_brfs,
          mcc_nns, f1_nns, accuracy_nns, mcc_nbs, f1_nbs, accuracy_nbs, 
          mcc_nb2, f1_nbs2, accuracy_nbs2))
metric_all_sampled

# Plot data
metric_plot_sampled <- metric_all_sampled %>%
ggplot(aes(Model, Value, fill = Metric)) +
geom_col(position = "dodge")
metric_plot_sampled

# ggsave(
#   "metricplot2.png",
#   plot = last_plot(),
#   device = "png",
#   path = "~/Downloads",
#   scale = 0.5,
#   width = 40,
#   height = 15,
#   units = "cm",
#   dpi = 600,
#   limitsize = TRUE,
#   bg = NULL
# )
```

# Supplimentary
## Variable importance

```{r rfvs, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
# variable importance using random forest

# load packages
library(varImp)
library(caret)
# Fit Random Forest model
set.seed(123)
importance_model <- randomForest(formula_train, data = preECMO, importance=TRUE) 

#Conditional=True, adjusts for correlations between predictors.
# ls(importance_model)
i_score <- randomForest::importance(importance_model)
i_score
varImpPlot(importance_model, sort = TRUE,  main = "Title", pch=16 ) # plot importance

# library(xtable)
# print(xtable(i_score, type = "latex"), file = "importanceRF.tex") # latex table

```



